When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-20 doc/paper/vpd_et_paper.tex
starting to clean up for publication

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and relative humidity drying ** RH drying \citep{Byrne_2013}. Increases in VPD increases the atmospheric demand of evapotranspiration (ET) \citep{Monteith_1965} but also stresses plant stomata \citep{Leuning_1990, MEDLYN_2011}.

This conflicting response between increased demand and higher stomatal stress leads to two potential but opposing perspectives on how ET adjusts to changes in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion \citep{Swann_2016}. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. Therefore, an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.

>>>>

Vapor pressure deficit (VPD) is expected to rise over continents in the future due to the combination of increased temperature and relative humidity drying ** RH drying? \citep{Byrne_2013}. Increases in VPD increase the atmospheric demand for evapotranspirated water \citep{Penman_1948, Monteith_1965}, but also stress plant stomata \citep{Leuning_1990, MEDLYN_2011}. These  conflicting responses, increased atmospheric demand and higher stomatal stress, leads to two possible responses of evapotraspiration (ET) to changes in VPD. The first, hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand with limited surface regulation though ecosystem resistance regualtion \citep{Swann_2016}. On the other hand, there is another perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. Therefore, an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET.
